---
title: "Cost-Effectiveness Applications"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Cost-Effectiveness Applications}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

# Numeric notation in this document
# With thanks to https://stackoverflow.com/questions/18965637/set-global-thousand-separator-on-knitr/18967590#18967590
knitr::knit_hooks$set(
    inline = function(x) {
        if(!is.numeric(x)){
            x
        } else if(x<100) {
            prettyNum(round(x, 3), big.mark=",")
        } else {
            prettyNum(round(x, 0), big.mark=",")
        }
    }
)
```

## Motivation

Cost-efffectiveness models evaluate the costs and health outcomes associated with a decision, such as treating a patient with a particular intervention, as compared with another. Such analyses depend on measuring the incremental costs and health outcomes between the different decisions, and considering which decision would be most efficient, given the decision-maker's economic perspective, time horizon, and willingness-to-pay for any improvement in health outcomes. Key to such analysis are the calculation of discounted (net) present values of costs and health outcomes associated with each decision option. Decisions about access to health care innovations are routinely made by health technology assessment (HTA) bodies on the basis of analyses in this form.

One of the problems with such analyses is that they are static in terms of pricing and uptake assumptions. Uptake is static because the analyses are conducted in respect of a single cohort of patients for whom the decision must be made today. In contrast, the decision-maker is likely to be making decisions on behalf of a gradual influx of patients for whom the decision is relevant. That influx may be dependent on disease prevalence, disease incidence, diagnosis patterns, progression of disease, prior treatment patterns - any number of the factors that health economists routinely consider in budget impact models, but not cost-effectiveness modeling.

Pricing is static because the analyses typically assume that costs will remain constant in real terms. When considering the cost-effectiveness of medicines, the acquisition cost of those medicines is a key component of the total cost. One thing we know about drug acquisition costs is that they do not remain constant in real terms. In the United States, companies often raise the Wholesale Acquisition Cost (WAC) prices of branded medicines once or twice a year, and this can be greater than inflation rates in the economy. After the exclusivity period of those branded medicines, generics or biosimilars may become available, and this wider supply can force down prices dramatically. In Europe and elsewhere, companies are often prevented - either directly in law or indirectly through HTA processes - from raising prices on branded medicines. Therefore assuming static prices is often a highly flawed assumption when using these analyses to make decisions about branded medicines, whose price may vary in future years - but absolutely not remain constant in real terms.

This vignette describes how to use *dynamicpv* to evaluate cost-effectiveness with and without dynamic pricing and dynamic uptake. It is intended to be read after `vignette("Dynamic Pricing")` and `vignette("Dynamic Uptake")`.

Before we begin, let us load the packages we need.

```{r setup, echo=TRUE, message=FALSE}
library(dplyr)
library(ggplot2)
library(lubridate)
library(flexsurv)
library(heemod)
library(dynamicpv)
```

## Assumptions

Before we start, we need to outline our assumptions. These concern:

- the decision problem we are modeling through a cost-effectiveness analysis
- dynamic patient uptake
- dynamic pricing

### Decision problem we are modeling

We wish to evaluate the cost-effectiveness, measured as incremental cost per QALY, of a new intervention compared to the standard of care (SoC). The model is a partitioned survival analysis typical in oncology with three health states: progression-free (PF), progressive disease (PD) and death, with additional assumptions as follows:

- The time horizon is 20 years, a timestep is 1 week, discounting is at 3% per year, with an effective date of calculation is 2025-06-01.
- For SoC, progression-free survival (PFS) is modeled as an exponential distribution with mean of 50 weeks; overall survival (OS) in weeks is modelled according to a lognormal distribution with mean and standard deviation on the log scale as 4 and 1 respectively.
- The effect of the new intervention relative to the SoC is a hazard ratio of 0.5 on PFS and 0.6 on OS.
- Health utility is 0.8 in the PF state and 0.6 in the PD state.
- Drug acquisition costs are $400 and $1500 per week for SoC and the new intervention respectively, incurred throughout a patient’s time in the PF state (treat to progression).
- Drug administration costs are $50 and $75 per week on treatment for SoC and the new intervention respectively.
- Adverse events occur in 0.08% and 0.10% of SoC and new intervention patients per week respectively, while on treatment, at an average cost of $2000 per event.
- Disease management costs are $80 per week in PF and $20 per week in PD.
- Patients receive the treatment of interest (new or SoC) until progression. Subsequent treatment costs are $1200 and $300 per week per patient while in the PD state for SoC and new intervention patients respectively.

We code the time constants, time horizon, discount rates and inflation rates first.

```{r constants}
# Time constants
days_in_year <- 365.25
days_in_week <- 7
cycle_years <- days_in_week / days_in_year # Duration of a one week cycle in years

# Time horizon (years) and number of cycles
thoz <- 20
Ncycles <- ceiling(thoz/cycle_years)

# Real discounting
disc_year <- 0.03 # Per year
disc_cycle <- (1+disc_year)^(cycle_years) - 1 # Per cycle

# Price inflation
infl_year <- 0.025 # Per year
infl_cycle <- (1+infl_year)^(cycle_years) - 1 # Per cycle

# Nominal discounting
nomdisc_year <- (1+disc_year)*(1+infl_year) - 1
nomdisc_cycle <- (1+nomdisc_year)^(cycle_years) - 1 # Per cycle
```

This model may then be coded in *heemod* as follows.

```{r heemod}
# State names
state_names = c(
  progression_free = "PF",
  progression = "PD",
  death = "Death"
  )

# PFS distribution for SoC with Exp() distribution and mean of 50 weeks
surv_pfs_soc <- heemod::define_surv_dist(
  distribution = "exp",
  rate = 1/50
)

# OS distribution for SoC with Lognorm() distribution, meanlog = 4.5, sdlog = 1
# This implies a mean of exp(4 + 0.5 * 1^2) = exp(4.5) = 90 weeks
surv_os_soc <- heemod::define_surv_dist(
  distribution = "lnorm",
  meanlog = 4,
  sdlog = 1
)

# PFS and OS distributions for new
surv_pfs_new <- heemod::apply_hr(surv_pfs_soc, hr=0.5)
surv_os_new <- heemod::apply_hr(surv_os_soc, hr=0.6)

# Define partitioned survival model, soc
psm_soc <- heemod::define_part_surv(
  pfs = surv_pfs_soc,
  os = surv_os_soc,
  terminal_state = FALSE,
  state_names = state_names
  )

# Define partitioned survival model, soc
psm_new <- heemod::define_part_surv(
  pfs = surv_pfs_new,
  os = surv_os_new,
  terminal_state = FALSE,
  state_names = state_names
  )

# Parameters
params <- heemod::define_parameters(
  # Discount rate
  disc = disc_cycle,
  # Disease management costs
  cman_pf = 80,
  cman_pd = 20,
  # Drug acquisition costs - the SoC regime only uses SoC drug, the New regime only uses New drug
  cdaq_soc = dispatch_strategy(
    soc = 400,
    new = 0
  ),
  cdaq_new = dispatch_strategy(
    soc = 0,
    new = 1500
  ),
  # Drug administration costs
  cadmin = dispatch_strategy(
    soc = 50,
    new = 75
  ),
  # Adverse event risks
  risk_ae = dispatch_strategy(
    soc = 0.08,
    new = 0.1
  ),
  # Adverse event average costs
  uc_ae = 2000,
  # Subsequent treatments
  csubs = dispatch_strategy(
    soc = 1200,
    new = 300
  ),
  # Health state utilities
  u_pf = 0.8,
  u_pd = 0.6,
)

# Define PF states
state_PF <- heemod::define_state(
  # Costs for the state
  cost_daq_soc = discount(cdaq_soc, disc_cycle),
  cost_daq_new = discount(cdaq_new, disc_cycle),
  cost_dadmin = discount(cadmin, disc_cycle),
  cost_dman = discount(cman_pf, disc_cycle),  
  cost_ae = risk_ae * uc_ae,
  cost_subs = 0,
  cost_total = cost_daq_soc + cost_daq_new + cost_dadmin + cost_dman + cost_ae + cost_subs,
  # Health utility, QALYs and life years
  pf_year = discount(cycle_years, disc_cycle),
  life_year = discount(cycle_years, disc_cycle),
  qaly = discount(cycle_years * u_pf, disc_cycle)
  )

# Define PD states
state_PD <- heemod::define_state(
  # Costs for the state
  cost_daq_soc = 0,
  cost_daq_new = 0,
  cost_dadmin = 0,
  cost_dman = discount(cman_pd, disc_cycle),  
  cost_ae = 0,
  cost_subs = discount(csubs, disc_cycle),
  cost_total = cost_daq_soc + cost_daq_new + cost_dadmin + cost_dman + cost_ae + cost_subs,
  # Health utility, QALYs and life years
  pf_year = 0,
  life_year = heemod::discount(cycle_years, disc_cycle),
  qaly = heemod::discount(cycle_years * u_pd, disc_cycle)
  )

# Define Death state
state_Death <- heemod::define_state(
  # Costs are zero
  cost_daq_soc = 0,
  cost_daq_new = 0,
  cost_dadmin = 0,
  cost_dman = 0,
  cost_ae = 0,
  cost_subs = 0,
  cost_total = cost_daq_soc + cost_daq_new + cost_dadmin + cost_dman + cost_ae + cost_subs,
  # Health outcomes are zero
  pf_year = 0,
  life_year = 0,
  qaly = 0,
)

# Define strategy for SoC
strat_soc <- heemod::define_strategy(
    transition = psm_soc,
    "PF" = state_PF,
    "PD" = state_PD,
    "Death" = state_Death
  )

# Define strategy for new
strat_new <- heemod::define_strategy(
  transition = psm_new,
  "PF" = state_PF,
  "PD" = state_PD,
  "Death" = state_Death
)

# Create heemod model
heemodel <- heemod::run_model(
  soc = strat_soc,
  new = strat_new,
  parameters = params,
  cycles = Ncycles,
  cost = cost_total,
  effect = qaly,
  init = c(1, 0, 0),
  method = "life-table"
)
```

### Dynamic pricing

Costs are assumed to increases in line with general inflation (2.5% per year), except for effects on drug acquisition costs due to LoEs. The date of calculation is 2025-09-01. The LoE for the SoC is assumed to occur first, at 2028-01-01, after which there is anticipated to be a 70% reduction in prices over one year. The new intervention has an LoE occuring three years later, at 2031-01-01, after which there would be a 50% reduction in prices over one year.

```{r dynpricing}
# Dates
# Date of calculation = 1 September 2025
doc <- lubridate::ymd("20250901")
# Date of LOE for SoC = 1 January 2028
loe_soc_start <- lubridate::ymd("20280101")
# Maturation of SoC prices by LOE + 1 year, i.e. = 1 January 2029
loe_soc_end <- lubridate::ymd("20290101") 
# Date of LOE for new treatment = 1 January 2031
loe_new_start <- lubridate::ymd("20310101")
# Maturation of new treatment prices by LOE + 1 year, i.e. = 1 January 2032
loe_new_end <- lubridate::ymd("20320101") 

# Effect of LoEs on prices once mature
loe_effect_soc <- 0.7
loe_effect_new <- 0.5

# Calculation of weeks since DoC for LoEs and price maturities
wk_start_soc <- floor((loe_soc_start-doc) / lubridate::dweeks(1))
wk_end_soc <- floor((loe_soc_end-doc) / lubridate::dweeks(1))
wk_start_new <- floor((loe_new_start-doc) / lubridate::dweeks(1))
wk_end_new <- floor((loe_new_end-doc) / lubridate::dweeks(1))

# Price maturity times
wk_mature_soc <- wk_end_soc - wk_start_soc
wk_mature_new <- wk_end_new - wk_start_new

# Create a tibble of price indices of length 2T, then pull out columns as needed
# We only need of length T for now, but need of length 2T for future calculations later
pricetib <- dplyr::tibble(
  model_time = 1:(2*Ncycles),
  model_year = model_time * cycle_years,
  static = 1,
  geninfl = (1 + infl_cycle)^(model_time - 1),
  loef_soc = pmin(pmax(model_time - wk_start_soc, 0), wk_mature_soc) / wk_mature_soc,
  loef_new = pmin(pmax(model_time - wk_start_new, 0), wk_mature_new) / wk_mature_new,
  dyn_soc = geninfl * (1 - loe_effect_soc * loef_soc),
  dyn_new = geninfl * (1 - loe_effect_new * loef_new)
)

# Price indices required for calculations
prices_oth <- pricetib$geninfl
prices_static <- pricetib$static
prices_dyn_soc <- pricetib$dyn_soc
prices_dyn_new <- pricetib$dyn_new
```

### Dynamic patient uptake

A disease incidence of 1 patient per timestep is assumed, with zero prevalence for simplicity. Under the decision problem to be evaluated, when the new intervention is
available, uptake is assumed to rise linearly from 0% to 100% after 2 years. In this way, uptake would be gradually increasing with time, accounting for disease epidemiology and the share of patients who receive the new intervention. 

```{r dynuptake}
# Time for uptake to occur
uptake_years <- 2

# Uptake vector for non-dynamic uptake
uptake_single <- c(1, rep(0, Ncycles-1))

# Uptake vector for dynamic uptake
uptake_weeks <- round(uptake_years / cycle_years)
share_multi <- c((1:uptake_weeks)/uptake_weeks, rep(1, Ncycles-uptake_weeks))
uptake_multi <- rep(1, Ncycles) * share_multi
```

## Using *dynamicpv*

### Without dynamic pricing or dynamic uptake

The conventional cost-effectiveness model is static.

```{r static1}
heemodel
```

Let's examine each payoff more closely.

```{r static2}
# Pull out the payoffs of interest from oncpsm
payoffs <- get_dynfields(
    heemodel = heemodel,
    payoffs = c("cost_daq_new", "cost_daq_soc", "cost_total", "qaly", "life_year"),
    discount = "disc"
    ) |>
    dplyr::mutate(
      model_years = model_time * cycle_years,
      # Derive costs other than drug acquisition, as at time zero
      cost_nondaq = cost_total - cost_daq_new - cost_daq_soc,
      # ... and at the start of each timestep
      cost_nondaq_rup = cost_total_rup - cost_daq_new_rup - cost_daq_soc_rup
    )

# Create and view dataset for SoC
hemout_soc <- payoffs |> dplyr::filter(int=="soc")
head(hemout_soc)

# Create and view dataset for new intervention
hemout_new <- payoffs |> dplyr::filter(int=="new")
head(hemout_new)
```

### Scenario 1: No dynamic uptake or pricing

```{r scen1}
#| output: false
#| echo: false

# SOC
s1_soc_othcost <- dynamicpv::dynpv(
    uptakes = uptake_single,
    payoffs = hemout_soc$cost_nondaq_rup,
    prices = prices_oth,
    discrate = nomdisc_cycle
    )$results$total
s1_soc_daqcost <- dynamicpv::dynpv(
    uptakes = uptake_single,
    payoffs = hemout_soc$cost_daq_soc_rup,
    prices = prices_static,
    discrate = disc_cycle
    )$results$total
s1_soc_cost <- s1_soc_daqcost + s1_soc_othcost
s1_soc_qaly <- dynamicpv::dynpv(
    uptakes = uptake_single,
    payoffs = hemout_soc$qaly_rup,
    prices = prices_static,
    discrate = disc_cycle
    )$results$total

# New intervention
s1_new_othcost <- dynamicpv::dynpv(
    uptakes = uptake_single,
    payoffs = hemout_new$cost_nondaq_rup,
    prices = prices_oth,
    discrate = nomdisc_cycle
    )$results$total
s1_new_daqcost <- dynamicpv::dynpv(
    uptakes = uptake_single,
    payoffs = hemout_new$cost_daq_new_rup,
    prices = prices_static,
    discrate = disc_cycle
    )$results$total
s1_new_cost <- s1_new_daqcost + s1_new_othcost
s1_new_qaly <- dynamicpv::dynpv(
    uptakes = uptake_single,
    payoffs = hemout_new$qaly_rup,
    prices = prices_static,
    discrate = disc_cycle
    )$results$total

# Incrementals
s1_icost <- s1_new_cost-s1_soc_cost
s1_iqaly <- s1_new_qaly-s1_soc_qaly
s1_icer <- s1_icost / s1_iqaly
```

These results show that the new intervention is associated with $`r s1_icost` incremental costs relative to the standard of care (including $`r s1_new_daqcost` of drug acquisition costs for the new intervention) and `r s1_iqaly` incremental QALYs. The cumulative ICER (incremental cost per QALY) is $`r s1_icer` per QALY at the 20 year time horizon.

### Scenario 2: Dynamic pricing, no dynamic uptake

```{r calc_scen2}
#| output: false
#| echo: false

# SOC
s2_soc_othcost <- s1_soc_othcost
s2_soc_daqcost <- dynamicpv::dynpv(
    uptakes = uptake_single,
    payoffs = hemout_soc$cost_daq_soc_rup,
    prices = prices_dyn_soc,
    discrate = nomdisc_cycle
    )$results$total
s2_soc_cost <- s2_soc_daqcost + s2_soc_othcost
s2_soc_qaly <- s1_soc_qaly

# New intervention
s2_new_othcost <- s1_new_othcost
s2_new_daqcost <- dynamicpv::dynpv(
    uptakes = uptake_single,
    payoffs = hemout_new$cost_daq_new_rup,
    prices = prices_dyn_new,
    discrate = nomdisc_cycle
    )$results$total
s2_new_cost <- s2_new_daqcost + s2_new_othcost
s2_new_qaly <- s1_new_qaly

# Incrementals
s2_icost <- s2_new_cost-s2_soc_cost
s2_iqaly <- s2_new_qaly-s2_soc_qaly
s2_icer <- s2_icost / s2_iqaly
```

Under scenario 2, the new intervention has an incremental cost-effectiveness of $`r s2_icer` per QALY (incremental costs of $`r s2_icost`, incremental QALYs of `r s2_iqaly`).

### Scenario 3: Dynamic uptake, not dynamic pricing

```{r calc_scen3}
#| output: false
#| echo: false

# SOC
s3_soc_othcost <- dynamicpv::dynpv(
    uptakes = uptake_multi,
    payoffs = hemout_soc$cost_nondaq_rup,
    prices = prices_oth,
    discrate = nomdisc_cycle
    )$results$total
s3_soc_daqcost <- dynamicpv::dynpv(
    uptakes = uptake_multi,
    payoffs = hemout_soc$cost_daq_soc_rup,
    prices = prices_static,
    discrate = disc_cycle
    )$results$total
s3_soc_cost <- s3_soc_daqcost + s3_soc_othcost
s3_soc_qaly <- dynamicpv::dynpv(
    uptakes = uptake_multi,
    payoffs = hemout_soc$qaly_rup,
    prices = prices_static,
    discrate = disc_cycle
    )$results$total

# New intervention
s3_new_othcost <- dynamicpv::dynpv(
    uptakes = uptake_multi,
    payoffs = hemout_new$cost_nondaq_rup,
    prices = prices_oth,
    discrate = nomdisc_cycle
    )$results$total
s3_new_daqcost <- dynamicpv::dynpv(
    uptakes = uptake_multi,
    payoffs = hemout_new$cost_daq_new_rup,
    prices = prices_static,
    discrate = disc_cycle
    )$results$total
s3_new_cost <- s3_new_daqcost + s3_new_othcost
s3_new_qaly <- dynamicpv::dynpv(
    uptakes = uptake_multi,
    payoffs = hemout_new$qaly_rup,
    prices = prices_static,
    discrate = disc_cycle
    )$results$total

# Incrementals
s3_icost <- s3_new_cost-s3_soc_cost
s3_iqaly <- s3_new_qaly-s3_soc_qaly
s3_icer <- s3_icost / s3_iqaly
```

Under scenario 3, the new intervention has an incremental cost-effectiveness of $`r s3_icer` per QALY (incremental costs of $`r s3_icost`, incremental QALYs of `r s3_iqaly` from a cohort comprising `r sum(uptake_multi)` patients).

### Scenario 4: Dynamic pricing and uptake

```{r calc_scen4}
#| output: false
#| echo: false

# SOC
s4_soc_othcost <- s3_soc_othcost
s4_soc_daqcost <- dynamicpv::dynpv(
    uptakes = uptake_multi,
    payoffs = hemout_soc$cost_daq_soc_rup,
    prices = prices_dyn_soc,
    discrate = nomdisc_cycle
    )$results$total
s4_soc_cost <- s4_soc_daqcost + s4_soc_othcost
s4_soc_qaly <- s3_soc_qaly

# New intervention
s4_new_othcost <- s3_new_othcost
s4_new_daqcost <- dynamicpv::dynpv(
    uptakes = uptake_multi,
    payoffs = hemout_new$cost_daq_new_rup,
    prices = prices_dyn_new,
    discrate = nomdisc_cycle
    )$results$total
s4_new_cost <- s4_new_daqcost + s4_new_othcost
s4_new_qaly <- s3_new_qaly

# Incrementals
s4_icost <- s4_new_cost-s4_soc_cost
s4_iqaly <- s4_new_qaly-s4_soc_qaly
s4_icer <- s4_icost / s4_iqaly
```

Under scenario 4, the new intervention has an incremental cost-effectiveness of $`r s4_icer` per QALY (incremental costs of $`r s4_icost`, incremental QALYs of `r s4_iqaly` from a cohort comprising `r sum(uptake_multi)` patients).

Total costs and QALYs for each scenario are summarized in @tbl-res1. The results above are skewed by the fact that some scenarios represent more than one patient cohort. Presenting results per patient, allows easier comparison between the scenarios.

|    |  | Scenario 1 | Scenario 2 | Scenario 3 | Scenario 4 |
|:---|:--|------------|------------|------------|------------|
| Dynamic pricing? || No | Yes | No | Yes |
| Dynamic uptake? || No | No | Yes | Yes |
| Cohort size || 1 | 1 | `r sum(uptake_multi)` | `r sum(uptake_multi)` |
| Total costs (cohort) | | ||||
| | New intervention | `r s1_new_cost` | `r s2_new_cost` | `r s3_new_cost` | `r s4_new_cost` |
| | Standard of care | `r s1_soc_cost` | `r s2_soc_cost` | `r s3_soc_cost` | `r s4_soc_cost` |
| | Incremental | `r s1_icost` | `r s2_icost` | `r s3_icost` | `r s4_icost` |
| Total costs (per patient) | | ||||
| | New intervention | `r s1_new_cost` | `r s2_new_cost` | `r s3_new_cost/sum(uptake_multi)` | `r s4_new_cost/sum(uptake_multi)` |
| | Standard of care | `r s1_soc_cost` | `r s2_soc_cost` | `r s3_soc_cost/sum(uptake_multi)` | `r s4_soc_cost/sum(uptake_multi)` |
| | Incremental | `r s1_icost` | `r s2_icost` | `r s3_icost/sum(uptake_multi)` | `r s4_icost/sum(uptake_multi)` |
| Total QALYs (cohort) | ||||
| | New intervention | `r s1_new_qaly` | `r s2_new_qaly` | `r s3_new_qaly` | `r s4_new_qaly` |
| | Standard of care | `r s1_soc_qaly` | `r s2_soc_qaly` | `r s3_soc_qaly` | `r s4_soc_qaly` |
| | Incremental | `r s1_iqaly` | `r s2_iqaly` | `r s3_iqaly` | `r s4_iqaly` |
| Total QALYs (per patient) | ||||
| | New intervention | `r s1_new_qaly` | `r s2_new_qaly` | `r s3_new_qaly/sum(uptake_multi)` | `r s4_new_qaly/sum(uptake_multi)` |
| | Standard of care | `r s1_soc_qaly` | `r s2_soc_qaly` | `r s3_soc_qaly/sum(uptake_multi)` | `r s4_soc_qaly/sum(uptake_multi)` |
| | Incremental | `r s1_iqaly` | `r s2_iqaly` | `r s3_iqaly/sum(uptake_multi)` | `r s4_iqaly/sum(uptake_multi)` |
| ICER | | `r s1_icer` | `r s2_icer` | `r s3_icer` | `r s4_icer` |

: Cost-effectiveness model results by scenario {#tbl-res1}

## Future single cohort ICER

The table above presents the cost-effectiveness results as of the date of calculation, `r doc`. However, it is interesting to explore how the ICER will change over time, given the expected evolution of prices. We use `dynamicpv::futurepv()` to calculate present values at future times. This function is a wrapper for `dynamicpv::dynpv()`. This is the single cohort ICER (no dynamic uptake) but with dynamic pricing, so corresponds with Scenario 2 from earlier.

First, we calculate the costs at each time of interest. Then we calculate the ICER, given the incremental QALYs we have already observed - and which are immune from pricing effects.

```{r future_calc}
# Times at which to plot ICER
gtimes <- round((0:(2*thoz))/cycle_years/2)
# SOC drug acquisition costs
gc_soc_daq <- gtimes |>
    purrr::map_vec(\(l) dynamicpv::futurepv(
        tzero = l,
        payoffs = hemout_soc$cost_daq_soc_rup,
        prices = prices_dyn_soc,
        discrate = nomdisc_cycle
    )$pv
)
# SOC other costs
gc_soc_oth <- gtimes |>
    purrr::map_vec(\(l) dynamicpv::futurepv(
        tzero = l,
        payoffs = hemout_soc$cost_nondaq_rup,
        prices = prices_oth,
        discrate = nomdisc_cycle
    )$pv
)
# New drug acquisition costs
gc_new_daq <- gtimes |>
    purrr::map_vec(\(l) dynamicpv::futurepv(
        tzero = l,
        payoffs = hemout_new$cost_daq_new_rup,
        prices = prices_dyn_new,
        discrate = nomdisc_cycle
    )$pv
)
# New other costs
gc_new_oth <- gtimes |>
    purrr::map_vec(\(l) dynamicpv::futurepv(
        tzero = l,
        payoffs = hemout_new$cost_nondaq_rup,
        prices = prices_oth,
        discrate = nomdisc_cycle
    )$pv
)
# Combine in a tibble
ds <- tibble::tibble(
    time_weeks = gtimes,
    time_years = time_weeks * cycle_years,
    evaldate = doc + time_weeks * 7,
    daqcost_new = gc_new_daq,
    daqcost_soc = gc_soc_daq,
    othcost_new = gc_new_oth,
    othcost_soc = gc_soc_oth,
    qaly_soc = s1_soc_qaly,
    qaly_new = s1_new_qaly
  ) |>
  dplyr::mutate(
    totcost_new = daqcost_new + othcost_new,
    totcost_soc = daqcost_soc + othcost_soc,
    icost = totcost_new-totcost_soc,
    iqaly = qaly_new-qaly_soc,
    icer = icost/iqaly
  )
```

We should check at this point that the ICER we are starting at (`r ds$icer[1]`) matches the scenario 2 ICER from earlier (`r s2_icer`).

```{r calc_future}
# Graph of ICER vs incident cohort time
g_icert <- ggplot2::ggplot(
    data = ds,
    aes(x=evaldate, y=icer)
    ) +
  geom_line() +
  labs(
    title = "",
    x = "Evaluation date",
    y = "ICER"
    ) +
  geom_hline(yintercept = ds$icer[1], linetype='dotted') +
  geom_vline(xintercept = loe_new_start, linetype='dashed') +
  geom_vline(xintercept = loe_soc_start, linetype='dashed') +
  theme_minimal() +
  scale_y_continuous(
    labels = scales::comma,
    limits=c(0, 150000)
    )
```

The following plot shows the nominal ICER calculated at different times, given the pricing, LoE and other assumptions. The horizontal dotted line confirms that the initial ICER matches the value from Scenario 2. The vertical dashed lines mark the timings of the LoEs of first the standard of care, and then the new treatment.

```{r graph_future, fig.width=9, fig.height=6, fig.cap="ICER over time", echo=FALSE}
g_icert
```

## Discussion

The findings for this example model are as follows:

- The effect of assuming dynamic pricing rather than flat pricing over time is shown in the difference between results of scenarios 1 and 2. Scenario 1 assumed no increases in drug acquisition costs, but inflationary increases in other costs. Scenario 2 additionally assumed drug prices of the new intervention and SoC comparator would be eroded due to losses of exclusivity. This has a material impact on costs per patient of the new intervention (reducing from $`r s1_new_cost` to $`r s2_new_cost`). Accordingly the ICER reduces from $`r s1_icer` to $`r s2_icer` per QALY.

- The effect of modeling uptake dynamically rather than just the current cohort of patients is shown in the difference between results of scenarios 1 and 3. Dynamic modeling of uptake has the effect of weighting the costs and QALYs per patient in scenario 3 towards a later cohort of patients, where the accumulation of costs and QALYs over the time remaining in the payer's 20 year time horizon will be less. The incremental costs per patient reduced from $`r s1_new_cost` in scenario 1 to $`r s3_new_cost/sum(uptake_multi)` in scenario 3, whereas the incremental QALYs reduced from `r s1_new_qaly` to `r s3_new_qaly/sum(uptake_multi)` per patient. Overall, the ICER increased slightly from $`r s1_icer` to $`r s3_icer` per QALY.

- Scenario 4 illustrates the effects of both dynamic pricing and cohort modeling. The results per patient are identical to scenario 3, except that the cost per patient of the new intervention reduces from $`r s3_new_cost/sum(uptake_multi)` to $`r s4_new_cost/sum(uptake_multi)`, due to the impact of dynamic pricing. The ICER in this case is reduced considerably to $`r s4_icer` per QALY.

