---
title: "Dynamic Uptake"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Dynamic Uptake}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

# Numeric notation in this document
# With thanks to https://stackoverflow.com/questions/18965637/set-global-thousand-separator-on-knitr/18967590#18967590
knitr::knit_hooks$set(
    inline = function(x) {
        if(!is.numeric(x)){
            x
        } else if(x<100) {
            prettyNum(round(x, 3), big.mark=",")
        } else {
            prettyNum(round(x, 0), big.mark=",")
        }
    }
)
```

## What is dynamic uptake?

Dynamic uptake refers to the modeling of multiple series of payoffs/cashflows over time, rather than the modeling of just one series of payoffs/cashflows at a time. This may arise, for example, in considering the treatment costs of multiple cohorts of patients beginning their treatment courses at different times. There are also analogies to 'run-off triangles' used by general insurance actuaries to model the emergence of insurance claims over time following claim events.

This package is designed to be able to calculate presente values with dynamic uptake/multiple cohorts. This can be coupled with considerations of dynamic pricing, as covered in `vignette("Dynamic Pricing")`.

## Using *dynamicpv*

First let us load the packages we will use for this vignette.

```{r setup}
library(dynamicpv)
```

Let us work with the same cashflow as in `vignette("Dynamic Pricing")`, but assume all costs/prices are constant in real terms. The discount rate is 3 \% per year, real.

```{r demo1}
# A simple cashflow
cashflow <- c(110, 120, 130, 140, 150)

# (Real) discount rate of 3\% per timestep (year)
disc <- 0.03

# Discounting factor - used for base R calculations later
vt1 <- (1 + disc)^(-1 * (0:4))
```

### Non-dynamic uptake

First let us consider non-dynamic uptake, as covered in `vignette("Dynamic Pricing")`.

| Time | Cashflow 1 | Discount factor    | Product |
|------|----------|--------------------|---------|
| $t$  | $c_t$    | $v^t = (1+i)^{1-t}$ |         |
| 1 | `r cashflow[1]` | `r vt1[1]` | `r cashflow[1] * vt1[1]` |
| 2 | `r cashflow[2]` | `r vt1[2]` | `r cashflow[2] * vt1[2]` |
| 3 | `r cashflow[3]` | `r vt1[3]` | `r cashflow[3] * vt1[3]` |
| 4 | `r cashflow[4]` | `r vt1[4]` | `r cashflow[4] * vt1[4]` |
| 5 | `r cashflow[5]` | `r vt1[5]` | `r cashflow[5] * vt1[5]` |
| Total | `r sum(cashflow)` | | `r sum(vt1 * cashflow)` | 

: Calculation of NPV from a simple cashflow and given discount rate

In this case, there is a single cashflow series starting at timestep 1. This time, we will use the `uptakes` argument for `dynamicpv::dynpv()` - although this is unnecessary since its default is also 1.

```{r uptake1}
# Uptake vector is simple
uptakes1 <- 1

# NPV calculation
pv1 <- dynpv(payoffs=cashflow, uptakes=uptakes1, discrate=disc)
pv1$results@total
```

The result is unchanged from the `vignette("Dynamic Pricing")`, where we also showed the function could be verified with some simple vector arithmetic.

### Dynamic uptake, constant in time (unweighted)

Suppose we have one new patient each year, and we wish to calculate the total NPV in a time horizon of 5 years. Now we have a payoff triangle as follows.

| Time | Cashflow 1 | Cashflow 2 | Cashflow 3 | Cashflow 4 | Cashflow 5 | Cashflow Sum | Discount factor | Product |
|------|----------|--------------------|---------|----|-----|----|----|----|
| $t$  | $c_t$    | $c_{t-1}$ | $c_{t-2}$ | $c_{t-3}$ | $c_{t-4}$ |   | $v^t = (1+i)^{1-t}$ |         |
| 1 | `r cashflow[1]` | - | - | - | - | `r cashflow[1]` |`r vt1[1]` | `r cashflow[1] * vt1[1]` |
| 2 | `r cashflow[2]` | `r cashflow[1]` | - | - | - | `r sum(cashflow[1:2])` |`r vt1[2]` | `r sum(cashflow[1:2]) * vt1[2]` |
| 3 | `r cashflow[3]` | `r cashflow[2]` |`r cashflow[1]` | - | - | `r sum(cashflow[1:3])` | `r vt1[3]` | `r sum(cashflow[1:3]) * vt1[3]` |
| 4 | `r cashflow[4]` | `r cashflow[3]` | `r cashflow[2]` |`r cashflow[1]` | - | `r sum(cashflow[1:4])` | `r vt1[4]` | `r sum(cashflow[1:4]) * vt1[4]` |
| 5 | `r cashflow[5]` | `r cashflow[4]` | `r cashflow[3]` | `r cashflow[2]` |`r cashflow[1]` | `r sum(cashflow[1:5])` | `r vt1[5]` | `r sum(cashflow[1:5]) * vt1[5]` |
| Total | `r sum(cashflow)` | | | | | | | `r sum(vt1 * cumsum(cashflow))` | 

: Calculation of NPV from a simple cashflow and given discount rate

With `dynamicpv::dynpv()`, we just update the `uptakes` argument.

```{r uptake2}
# Uptake vector is simple
uptakes2 <- rep(1, 5)
uptakes2

# NPV calculation
pv2 <- dynpv(payoffs=cashflow, uptakes=uptakes2, discrate=disc)
pv2$results
```

There are `r sum(uptakes2)` patients with a total NPV of `r pv2$results@total`, which equates to `r pv2$results@mean` on average per patient. This can also be calculated in base R.

```{r verify2}
sum(vt1 * cumsum(cashflow))
```

### Dynamic uptake, not constant in time

In general, uptake will not be constant in time. Due to factors such as varying epidemiology (prevalence and incidence), prior treatment pathways (patient testing, patient diagnostics, prior treatment usage), as well as the specific uptake of the treatment of interest (patient share), uptake will also be highly variable over time. In this setting, the total and mean NPVs must reflect weightings applicable to each cashflow.

For example, suppose the number of patients receiving treatment increases by one each year. The weighting given to cashflow 1 would then be $1/(1+2+3+4+5)=6.67$ \%.

```{r uptake3}
# Uptake vector is simple
uptakes3 <- 1:5

# NPV calculation
pv3 <- dynpv(payoffs=cashflow, uptakes=uptakes3, discrate=disc)
pv3$results
```

There are `r sum(uptakes3)` patients with a total NPV of `r pv3$results@total`, which equates to `r pv3$results@mean` on average per patient. Verifying this result in base R also becomes more complicated.

```{r verify3}
# Verifying total NPV is now more complicated
checkpv <- rep(0, 5)
for (i in 1:5) {
  checkpv[i] <- sum(cashflow[1:i] * uptakes3[i:1] * vt1[i])
}
sum(checkpv)
```

### Dynamic pricing and uptake

The package becomes most powerful when considering both dynamic uptake and dynamic pricing. A simple call to `dynamicpv::dynpv()` replaces what would be rather more complicated with base functions, even in this toy example.

```{r uptake4}
# Price index
pinfl <- 0.01
pindex <- (1+pinfl)^(0:4)
pindex[4:5] <- 0.5

# Nominal discount rate
nomdisc <- (1+disc)*(1+pinfl)-1

# NPV calculation
pv4 <- dynpv(payoffs=cashflow, uptakes=uptakes3, prices=pindex, discrate=nomdisc)
pv4$results
```

There are `r sum(uptakes3)` patients with a total NPV of `r pv4$results@total`, which equates to `r pv4$results@mean` on average per patient.

## Summary

- Dynamic uptake occurs if the same cashflow series may begin at several different times, as can occur when modeling costs of treatment to populations over time (where uptake of that treatment varies in time) or the run-off of insurance claims. This can complicate calculations of NPV.
- The `dynamicpv::dynpv()` function can be used to calculate NPVs for arbitrary vectors of cashflows, at a given discount rate and a given expected uptake over time.
- In this way, the function can accommodate both dynamic uptake and dynamic pricing, as seen in the `vignette("Dynamic Pricing")`.